\documentclass[../cellseek_paper.tex]{subfiles}

\begin{document}

\begin{abstract}
  Cell tracking in microscopy videos remains a bottleneck in quantitative biology due to the complexity of existing tools that require extensive parameter tuning and domain expertise. While powerful platforms like TrackMate offer sophisticated algorithms, their numerous options and steep learning curves limit accessibility for many biologists. We present CellSeek, a streamlined cell tracking pipeline that leverages foundation models to eliminate the parameter optimization burden while maintaining state-of-the-art performance. Our approach integrates Cellpose-SAM, a specialized adaptation of the Segment Anything Model derived from the Cellpose framework for cellular segmentation, with an adapted version of Cutie, a video object segmentation network modified specifically for cell tracking by removing long-term memory and focusing on last-frame temporal context. This adaptation recognizes that cells within populations exhibit highly similar morphological features, making appearance-based long-term memory counterproductive. Crucially, CellSeek includes an intuitive graphical interface that transforms complex computer vision workflows into a simple point-and-click experience. We demonstrate that CellSeek achieves comparable accuracy to expert-tuned TrackMate configurations across diverse microscopy datasets while reducing analysis time from hours to minutes and requiring no specialized training. By democratizing access to cutting-edge tracking algorithms, CellSeek represents a paradigm shift toward user-friendly, foundation model-powered tools for quantitative biology.
\end{abstract}

\end{document}
